### 2025-03-12 paper/sections/experiments.qmd
Add first part of discussion (Section 5.1)

### Impact of hyperparameter settings
We extensively test the impact of three types of hyperparameters on the proposed training regime. Our complete results are available in the technical appendix; this section focuses on the main findings.

***Hyperparameters of the CE generators.*** First, we observe that CT is highly sensitive to hyperparameter settings but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time. Second, we note that the choice of a CE generator has a major impact on the results. For example, *REVISE* tends to perform the worst, most likely because it attempts to explain the decisions of a VAE that is an imperfect surrogate of the original model. Third, increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in any iteration. Fourth, the impact of $\tau$, the required decision threshold is more difficult to predict. On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes.

***Hyperparameters for penalties*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values for @eq-obj penalties, this effect is not very pronounced.

>>>>

### Impact of hyperparameter settings {#sec-hyperparameters}
We extensively test the impact of three types of hyperparameters on the proposed training regime. Our complete results are available in the technical appendix; this section focuses on the main findings.

***Hyperparameters of the CE generators.*** First, we observe that CT is highly sensitive to hyperparameter settings but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time. Second, we note that the choice of a CE generator has a major impact on the results. For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful]. Third, increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch. Fourth, the impact of $\tau$, the required decision threshold is more difficult to predict. On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes. In fact, we have generally found that a choice of $\tau=0.5$ leads to optimal results because it is associated with high proportions of mature counterfactuals.

***Hyperparameters for penalties.*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values for any of the other penalties in @eq-obj, this effect is not very pronounced.

### 2025-03-12 paper/sections/intro.qmd

Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning the representations directly from data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Modern advances in computing have made it possible to provide such models with ever growing degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models. Unfortunately, in doing so, the models learn increasingly complex and highly sensitive representations that we can no longer easily interpret.

@@

Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so called "lottery tickets" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success]. Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data. Thus, looking at the predictive performance, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two related challenges. Firstly, there is no dependable way to verify if such complex representations correspond to meaningful and plausible explanations. Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.

>>>>

Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning representations directly from data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Modern advances in computing have made it possible to provide such models with ever-growing degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models. Unfortunately, in doing so, models learn increasingly complex and highly sensitive representations that we can no longer easily interpret.

@@

Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success]. Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data. Thus, looking at the predictive performance, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two related challenges. Firstly, there is no dependable way to verify if such complex representations correspond to meaningful and plausible explanations. Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.

### 2025-03-12 paper/sections/method.qmd

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual. For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.]

>>>>

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual. For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see supplementary appendix.]

<!-- @sec-app-ce -->
